---
title: "Ma2"
author: "Sigbjørn Fjelland"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\underline{Various packages in use:}
```{r}
set.seed(98765432)
library(igraph)
library(expm) 
```

w -> wide
t -> tilde
h -> hat
m -> matrix


\newpage
\underline{problem 2.1}\newline

a)
```{r}
P <- matrix(nrow = 5, ncol = 5, dimnames = list(c("0","1","2","3","4"),
                                                c("0","1","2","3","4")))
P[1,] <- c(0.000, 0.000, 0.498, 0.000, 0.502)
P[2,] <- c(0.000, 0.527, 0.366, 0.107, 0.000)
P[3,] <- c(0.451, 0.549, 0.000, 0.000, 0.000)
P[4,] <- c(0.000, 0.474, 0.114, 0.412, 0.000)
P[5,] <- c(0.328, 0.249, 0.000, 0.000, 0.423)

graph_P <- graph_from_adjacency_matrix(P, weighted=TRUE,diag=TRUE)
plot(graph_P)
```
\newpage
\underline{problem 2.1}\newline

b)
```{r}
irreducibility <- function(P, M){
            P.new <- P
            for(i in 2:M){
                Pk <- P %^% i
                P.new <- P.new + Pk
            }
            
            P.bar <- (1/M)*P.new
            
            if(all(P.bar > 0)){
                print("Transition Matrix is Irreducible")
              
            }
            else{
                print("YTransition Matrix is Reducible")
            }
            
}

irreducibility(P,5)
```
\newpage
\underline{problem 2.1}\newline

c) 
```{r}
K = 50
M = 5
p0 <- as.vector(c(0.013, 0.214, 0.197, 0.375, 0.201))

K <- 50
pk <- matrix(nrow = K, ncol = M)
for (i in 1:K) {
    pk[i, ] <- p0 %*% (P%^%i)
}


plot(x = 1:K, y = pk[, 1], type = "l", col = 1, xlab = "t", ylab = "p",
ylim = c(min(pk), max(pk)))

lines(x = 1:K, y = pk[, 2], type = "l", col = 2)
lines(x = 1:K, y = pk[, 3], type = "l", col = 3)
lines(x = 1:K, y = pk[, 4], type = "l", col = 4)
lines(x = 1:K, y = pk[, 5], type = "l", col = 5)


```
\newpage
\underline{problem 2.1}\newline

d)
Stationary distribution by computing eigenvalues of P
```{r}
pi <- eigen(t(P))$vectors[,1]/sum(eigen(t(P))$vectors[,1])

pi_confirm <- pi %*% P 

print(pi_confirm - pi)
```


e)
Stationary distribution by iterting 50 times: $\pi=\pi\cdot P^{50}$
```{r}
pi_converged <- p0 %*% (P%^%50)

```


As we see the difference is small
```{r}
print(pi-pi_converged)
```

\newpage
\underline{problem 2.2}\newline
a)
function for simulating large samples
```{r}
simulMarkov<-function(p0,mP,n){
M<-dim(mP)[1]
X<-rep(0,n)
X[1]<-sample(M,1,prob=p0)
for(t in 2:n) X[t]<-sample(M,1,prob=mP[X[t-1],])
return(X)
}
```


```{r}
n<-100
X<-rep(0,n)
X[1]<-sample(M,1,prob=p0)
for(t in 2:n) X[t]<-sample(M,1,prob=P[X[t-1],])
```

\underline{problem 2.2}\newline
b)


```{r}
plot(x = 1:n, y = X, main = "Simulation of markov Chain based on (P, p0)", 
     xlab = "n (simulated timestep)", ylab = "X(t)", type = "S")
```


##problem 2.2
\newline
c)

```{r}

iz<-4
S_all<-which(X==iz)
Tn<-length(S_all)

Tn


```
\underline{problem 2.2}\newline
d)

```{r}
tau<-S_all[2:Tn]-S_all[1:(Tn-1)]

plot(x = 1:n, y = X, main = "Simulation of markov Chain based on (P, p0)", 
     xlab = "n (simulated timestep)", ylab = "X(t)", type = "S")

abline(v = tau, col="red", lwd=3, lty=2)
```

\underline{problem 2.3}\newline
a)


```{r}
#set.seed(1234)
X.10000 <- simulMarkov(p0=p0, mP=P,n=10000)

iz.10000 = 4 #X.10000[1]
S_all<-which(X.10000 == iz.10000)

Tn.10000<-length(S_all)

Tn.10000
```

b)

```{r}
tau.10000<-S_all[2:Tn.10000]-S_all[1:(Tn.10000-1)]

f_i0<-c(1:100)
for(j in 1:100){
f_i0[j]<-(1/10000)*length(which(tau.10000==j))
}


barplot(f_i0, xlab = "f_i0")
```

c)
```{r}
set.seed(123)
i0 <- 4
Tn.1 <- length(which(simulMarkov(p0 = p0, mP=P, n=10000) == i0))
tau.1 <- S_all[2:Tn.1] - S_all[1:(Tn.1 - 1)]
K <- 100 

f_hat <- numeric(K)

for (s in 1:K) {
  
    f_hat[s] <- length(which(tau.1 == s))
  }

  f_hat <- f_hat / (Tn.1 - 1)


barplot(f_hat, xlab = "f_hat")
```




  i)
```{r}
mu_hat_i0 <- (1/(Tn.1-1))*sum(tau.1[-1]) #AKA tou_bar
mu_hat_i0
```

ii)

```{r}
mu_tilde_i0 <- sum((1:K) * f_hat)
mu_tilde_i0 
```

iii)

```{r}
pi_i0 <- pi_converged[i0]
pi_i0
```

```{r}
rbind("1 / pi_i0" = 1 / pi_i0,
"mu_hat_i0" = mu_hat_i0,
"mu_tilde_i0" = mu_tilde_i0)


```

d)

```{r}

n = 10000
m = n -1
X = simulMarkov(p0 = p0, mP=P, n=n)


wtpi <- whpi <- numeric(5)
    for(i in 1:5){
          wtpi[i]<-m^(-1)*length((which(X==i)))
    }

wtpi
sum(wtpi)

```

e)


Emprircal Transition matrix:


```{r}
P_hat <- matrix(nrow = 5, ncol = 5, 0)
wtpi <-c()
whpi <-c()
for (t in 1:(length(X) - 1)) P_hat[X[t], 
              X[t + 1]] <- P_hat[X[t], X[t + 1]] + 1

  for (i in 1:5) P_hat[i, ] <- P_hat[i, ] / sum(P_hat[i, ])
  
P_hat
```

\underline{problem 2.4}\newline

we apply the amse matrix as in 2.3 (e)

a)
```{r}
P_hat
```

comparing by subtracting $P$ from $\hat{P}$
b)

```{r}
P-P_hat
```
as we see the difference is about two to thre decimals


